---
title: "Bike Sharing Program: Rental Demand Prediction from Weather"
author: "MattDSquared (Matthew D. Davis)"
date: "Monday, May 25, 2015"
output: 
    html_document:
        keep_md: true
---

_From the competition host, Kaggle.com_

Bike sharing systems are a means of renting bicycles where the process of obtaining membership, rental, and bike return is automated via a network of kiosk locations throughout a city. Using these systems, people are able rent a bike from a one location and return it to a different place on an as-needed basis. Currently, there are over 500 bike-sharing programs around the world.

The data generated by these systems makes them attractive for researchers because the duration of travel, departure location, arrival location, and time elapsed is explicitly recorded. Bike sharing systems therefore function as a sensor network, which can be used for studying mobility in a city. In this competition, participants are asked to combine historical usage patterns with weather data in order to forecast bike rental demand in the Capital Bikeshare program in Washington, D.C.

###Acknowledgements

Kaggle is hosting this competition for the machine learning community to use for fun and practice. This dataset was provided by Hadi Fanaee Tork using data from Capital Bikeshare. We also thank the UCI machine learning repository for hosting the dataset. If you use the problem in publication, please cite:

Fanaee-T, Hadi, and Gama, Joao, Event labeling combining ensemble detectors and background knowledge, Progress in Artificial Intelligence (2013): pp. 1-15, Springer Berlin Heidelberg.

## Project Setup

```{r setup, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggdendro)
library(lubridate)
library(plyr); library(dplyr)
source("myplclust.R")

sessionInfo()

setwd("~/../datascience/BikeSharingDemand")
graphics.off()
windows.options(width = 12, height = 8, xpinch=96, ypinch=96, xpos=0, ypos=0)
```

## Data Processing

Data is made available at https://www.kaggle.com/c/bike-sharing-demand/data. A Kaggle account is required. This needs to be downloaded manually (it's behind a login, which just complicates automated download). Alternatively, forking the BikeSharingDemand repo at https://github.com/MattDSquared/BikeSharingDemand will include this data set. 

```{r download, warning=FALSE}
dir.create("data")
train.filepath <- "data/train.csv"
test.filepath <- "data/test.csv"

# download train and test data sets from 
# https://www.kaggle.com/c/bike-sharing-demand/data
# place train.csv and test.csv into data subdirectory
# data obtained 2015-05-25
```

Load the data into train and test data sets.

```{r load}
train <- read.csv(train.filepath, stringsAsFactors=FALSE)
test <- read.csv(test.filepath, stringsAsFactors=FALSE)
```

Text data is cleaned up and simplified. The weather labels are a bit verbose, see the [competition data](https://www.kaggle.com/c/bike-sharing-demand/data) page for formal weather descriptions.

```{r clean.factorlabels}
weatherlabels <- c("nice weather", "cloudy/misty", "light weather",
                   "heavy weather")
seasons <- c("spring", "summer", "fall", "winter")
daysoftheweek <- c("Monday","Tuesday","Wednesday","Thursday","Friday",
                   "Saturday","Sunday")
train <- mutate(train, 
                datetime = ymd_hms(datetime),
                season = factor(season, levels=1:4, labels=seasons),
                holiday = as.logical(holiday),
                workingday = factor(workingday, levels=c(1,0), 
                                    labels=c("Workday","Holiday/Weekend")),
                weather = factor(weather,levels=1:4, labels=weatherlabels),
                dayofweek = factor(weekdays(datetime), levels=daysoftheweek),
                timeofday = hour(datetime))
#test <- mutate(test, ...)
#   TODO: fill out test data later once have solid train dataset

```

Data only shows a single heavy weather point. In the midst of several light weather points. For graphing purposes, replace with light weather. Implicit assumption: bike rentals during heavy weather will be negligeable. 

```{r clean.heavyweather}
print(which(train$weather == "heavy weather"))
print(which(test$weather == "heavy weather"))
print(train[which(train$weather == "heavy weather")+(-5):6,c("datetime","weather", "temp", "count")])

## interpolate heavy weather as light weather for graphing purposes
train$weather[train$weather == "heavy weather"] = "light weather"
test$weather[test$weather == "heavy weather"] = "light weather"
```

This should be explored further. Is this data set applicable to the DC area, since it essentially doesn't include heavy weather data which does occur in the DC area. 

## What is atemp?

The atemp variable in the data set is the "feels like" temperature based on weather factors. The plot below illustrates this relationship.

```{r atemp.plot}
## atemp vs. temperature, weather
gg <- ggplot(train, aes(temp, atemp, color=humidity, size=windspeed)) +
    geom_point() +
    scale_colour_gradient(low="blue",high="red") + 
    labs(title="Effects on 'Feels Like' Tempurature") +
    labs(x="tempurature (°C)") +
    labs(y="'feels like' tempurature (°C)")
windows()
print(gg)
```

This shows the effects of windchill below about 15°C and humidity above about 24°C. Note the data contains few points below 5°C, which could be an issue for the DC area. 

Also note the outliers to the lower right of the main grouping. These `r nrow(filter(train, atemp < 15, temp > 24))` data all occur on a single day in august.
Note the fixed atemp value of `r filter(train, atemp < 15, temp > 24)$atemp[1]`. For simplicity, assume for these values atemp = temp.

```{r atemp.outliers}
select(filter(train, atemp < 15, temp > 24), datetime, weather, temp, atemp, humidity, windspeed)

# replace atemp with temp for outliers
train <- mutate(train, atemp=ifelse((atemp < 15) & (temp > 24), temp, atemp))
test <- mutate(test, atemp=ifelse((atemp < 15) & (temp > 24), temp, atemp))
```

It appears wind and humidity are already packaged into the 'feels like' tempurature value. The following analysis neglects temp, wind, and humidity in favor of using atemp. This leaves the weather label, time (hour of the day, day of the week), and holiday label. 

## Primary demand drivers

### Effect of Time of day, weather, and working/non-working days

The below plot shows the effect of time, weather, and work/nonwork days on daily rental counts. The initial idea for this plot came from [this kaggle script post](https://www.kaggle.com/users/993/ben-hamner/bike-sharing-demand/bike-rentals-by-time-and-temperature) by Ben Hamner, which has been greatly expanded here. 

```{r demand.daily, fig.width=8, fig.height=6}
colors.tempurature <- c("#5e4fa2", "#3288bd", "#66c2a5", "#abdda4", "#e6f598",
                        "#fee08b", "#fdae61", "#f46d43", "#d53e4f", "#9e0142")

gg <- ggplot(train, aes(timeofday, count, color=9/5*atemp+32)) +
    facet_grid(workingday ~ weather) +
    geom_point() +
    geom_smooth() +
    theme(plot.title = element_text(size = rel(1.5))) +
    labs(title="Daily Bike Rental Demand \nPer Time of Day, Work/Nonwork day, and Weather") + 
    labs(x="Hour of Day") + 
    labs(y="Bike Rentals Initiated per Hour") +
    scale_colour_gradientn("Temp (°F)", colours=colors.tempurature)
windows()
print(gg)
```

Features to note:

* Sinusoidal shape of demand with respect to hour of the day.
* Weather does not greatly effect the the shape of the demand curve with respect to time of day, just the magnitude and density. 
* Weather does seem to effect the average temperature of the data.
* Dependance on demand to atemp, with peak demand in the 80-100°F range. 
* Distincly differing shape between a work and non-work day, with 
* Peak demand occurs in the afternoon of a warm, nice-weather, work-day.

### Principal component analysis (PCA)

PCA Analysis here.

## Model Development

Using a staged model development approach which attempts to remove the effect of each variable in isolated steps. 

### temperature model

### Work vs. Nonwork day

### Weather effect

### Time of day (final model)

## Future Work

Items remaining for future work:

* Include measured tempurature, wind, and humidity in the analysis. 
* Determine if data set is applicable to the DC weather patterns. Data appears to be from much more temperate climate which could greatly effect weather-based behavior. Theory: residents get accustomed to their "average" weather condition, adjusting outdoor activity accordingly. 